# Week 6. Analysis of longitudinal data

## Preparations

```{r warning=FALSE, message=FALSE}
# Set working directory
#setwd("Z:/Jatko-opinnot/IODS/IODS-project-1")
setwd("C:/Users/sirja/Documents/IODS-project-1")

# Access needed libraries
library("tidyverse")
library("lme4")
library("gridExtra")
```


```{r}
# load the data
BPRSL <- read.csv("data/BPRSL.txt", row.names = 1)
RATSL <- read.csv("data/RATSL.txt", row.names = 1)

# explore the dataset
str(BPRSL)
str(RATSL)
```

R doesn't recognise the factorised variables made in the meet_and_repeat.R script, so lets factorise them again.

```{r}
# Convert the categorical variables to factors
BPRSL$treatment <- factor(BPRSL$treatment)
BPRSL$subject <- factor(BPRSL$subject)

RATSL$ID <- factor(RATSL$ID)
RATSL$Group <- factor(RATSL$Group)

str(BPRSL)
str(RATSL)
```

Now data should be ready for analyses.


## RATSL data

Draw a line plot of the rat weights in diet groups 1-3 
```{r, fig.height=2.7, fig.width=7}
ggplot(RATSL, aes(x = Time, y = Weight, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(RATSL$Weight), max(RATSL$Weight)))

```

High start weight shows as high end weight, so data should be standardized.

```{r, fig.height=2.7, fig.width=7}
RATSL <- RATSL %>%
  group_by(Time) %>%
  mutate(stdweight = (Weight-mean(Weight))/sd(Weight)) %>%
  ungroup()

ggplot(RATSL, aes(x = Time, y = stdweight, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(RATSL$stdweight), max(RATSL$stdweight)), name = "standardized bprs")
```

Now let's see the response for the diets.

```{r, fig.height=5, fig.width=7}
n <- RATSL$Time %>% unique() %>% length()
RATS_trmt <- RATSL %>% group_by(Group, Time) %>% summarize(mean = mean(Weight), se = sd(Weight)/sqrt(n))
glimpse(RATS_trmt)

ggplot(RATSL, aes(x= as.factor(Time), y=Weight, fill=Group)) +
  geom_boxplot() +
  theme(legend.position = c(0.8,0.4), panel.grid = element_blank(), panel.background = element_blank(), panel.border = element_rect(fill = FALSE), legend.key = element_rect(fill=FALSE)) +
  scale_y_continuous(name = "mean(Weight) +/- se(Weight)") + 
  scale_x_discrete(name = "Time") + 
  scale_fill_manual(values=c("#e999ab", "#46ab9f", "#3e93c2")) +
  ggtitle("Rat weight on different diets over time")
```

Rats on diet 1 weigh less than rats on diet 2 and 3, on average.


```{r, fig.height=4, fig.width=4.5}
RATSL8 <- RATSL %>% 
  group_by(Group, ID) %>%
  summarize(mean = mean(Weight)) %>%
  ungroup()

ggplot(RATSL8, aes(x=Group, y=mean)) + 
  geom_boxplot() + 
  stat_summary(fun.y = "mean", geom = "point", shape=23, size=4, fill = "white") + 
  theme(legend.position = c(0.8,0.8), panel.grid = element_blank(), panel.background = element_blank(), panel.border = element_rect(fill = FALSE)) +
  scale_y_continuous(name = "mean(Weight) per group")
```

There is one outlier in each group, but in group 2 it differs a lot from the mean. Let's remove it and draw the boxplot again.

```{r, fig.height=4, fig.width=4.5}
RATSL8S1 <- filter(RATSL8, (Group=="2" & mean<500) | Group!="2")

ggplot(RATSL8S1, aes(x = Group, y = mean)) +
  geom_boxplot() +
  theme(legend.position = c(0.8,0.8), panel.grid = element_blank(), panel.background = element_blank(), panel.border = element_rect(fill = FALSE)) +
  stat_summary(fun.y = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(Weight) per group")
```


## BPRSL data

Let's draw a lineplot for an overview of the data.
```{r, fig.height=5, fig.width=5}
ggplot(BPRSL, aes(x = week, y = bprs, fill=subject, color = treatment)) +
  geom_line(aes(linetype = treatment)) +
  theme(legend.position = c(0.8,0.8), panel.grid = element_blank(), panel.background = element_blank(),
        panel.border = element_rect(fill = FALSE), legend.key = element_rect(fill=FALSE)) + 
  scale_x_continuous(name = "Time (weeks)", breaks = seq(0, 8, 2), expand=c(0,0)) + 
  scale_y_continuous(name = "bprss") +
  scale_color_manual(values=c("#e999ab", "#46ab9f")) +
  ggtitle("Symptoms in patients over time")
```

Just by eyeballing it seems that with treatment 1 the symptoms have decreased within time, whereas with treatment 2 there is more variation towards the end of the monitoring period.

Let's fit a Random Intercept Model.

```{r}
BPRS_ref <- lmer(bprs ~ week + treatment + (1 | subject), data = BPRSL, REML = FALSE)

summary(BPRS_ref)
```

Now let's fit a random intercept and random slope model and compare it to the random intercept model.
```{r}
BPRS_ref1 <- lmer(bprs ~ week + treatment + (week | subject), data = BPRSL, REML = FALSE)

summary(BPRS_ref1)

anova(BPRS_ref, BPRS_ref1)
```


Let's try interaction between week and treatment and compare again.
```{r}
BPRS_ref2 <- lmer(bprs ~ week * treatment + (week | subject), data = BPRSL, REML = FALSE)

summary(BPRS_ref2)

anova(BPRS_ref2, BPRS_ref1)
```

Based on the chi-squared statistics and p-value of the likelihood ratio test, random intercept and random slope model is the best. Let's plot:

```{r, fig.width=7, fig.height=4}
p1 <- ggplot(BPRSL, aes(x = week, y = bprs, fill = subject, color = treatment)) +
  geom_line(aes(linetype = treatment)) +
  scale_x_continuous(name = "Time (weeks)", breaks = seq(0, 8, 2)) +
  scale_y_continuous(name = "Observed bprs") +
    scale_color_manual(values=c("#e999ab", "#46ab9f")) +
    ggtitle("Symptoms in patients over time") +
  theme(panel.grid = element_blank(), panel.background = element_blank(),
        panel.border = element_rect(fill = FALSE), legend.key = element_rect(fill=FALSE), legend.position = "bottom",)
Fitted <- fitted(BPRS_ref1)
BPRSL <- mutate(BPRSL, fitted=Fitted)

p2 <- ggplot(BPRSL, aes(x = week, y = fitted, fill = subject, color = treatment)) +
  geom_line(aes(linetype = treatment)) +
  scale_x_continuous(name = "Time (weeks)", breaks = seq(0, 8, 2)) +
  scale_y_continuous(name = "Fitted bprs") +
  scale_color_manual(values=c("#e999ab", "#46ab9f")) +
    ggtitle(" ") +
  theme(panel.grid = element_blank(), panel.background = element_blank(),
        panel.border = element_rect(fill = FALSE), legend.key = element_rect(fill=FALSE), legend.position = "bottom")


# p1 and p2 side by side
grid.arrange(p1, p2, ncol=2)
```

Based on the fitted model, it looks like the treatments are after all equally effective in reducing the symptoms.
