# Week 2: Analysis excercise

## Preparations
```{r}
# Set working directory
setwd("Z:/Jatko-opinnot/IODS/IODS-project-1")

# Access needed libraries
library(ggplot2)
library(GGally)
```

## Exercise 1
```{r}
# Read the learning2014 data
learning2014 <- read.csv("data/learning2014.csv")

# Explore the data
str(learning2014)
dim(learning2014)

```

## Exercise 2
```{r}
# create a plot matrix with ggpairs()
p <- ggpairs(learning2014, mapping = aes(col = gender, alpha = 0.3), lower = list(combo = wrap("facethist", bins = 20)))

# draw the plot
p
```





***
> [Here](https://www.mv.helsinki.fi/home/kvehkala/JYTmooc/JYTOPKYS3-meta.txt) you can find more information on the data. 



## Exercise 3
```{r}
# create a regression model with three explanatory variables
reg_model <- lm(points ~ attitude + stra + age, data = learning2014)

# print out a summary of the model
summary(reg_model)

# Try with other variables
reg_model2 <- lm(points ~ stra + age + deep, data = learning2014)

# print out a summary of the model
summary(reg_model2)
```




## Exercise 4
By trying different variable combinations in the model, it could be seen that attitude had significant relationship with the points, whenever it was included in the model. In addtion, strategic learning had a significant correlation with the points, when attitude was not included in the model. Age had some correlation with the points. but it was not significant.

## Exercise 5
```{r}
# draw diagnostic plots : Residuals vs Fitted values, Normal QQ-plot and Residuals vs Leverage
plot(reg_model, which = c(1,2,5), par(mfrow = c(2,2)))

```
